Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-07-23 surfaces-write-up/surface-write-up.tex
Changes to hirzebruch section

\begin{lem}
\label{lem:hirz-generators}
$R_D$ is generated in degree at most
\[
	\sigma = \sum_{i \in \Te(D)} \gcd(\bida_i, \bidb_i) \ell_i + \sum_{\substack
	{i \in \Tp(D) \\ j\in \Tm(D)}} (\bida_i \bidb_j - \bida_j \bidb_i)
	\ell	_{i,j}.
\]
\end{lem}

@@

\noindent
for some $c_i \ge - \alpha_i d$ such that $\sum_{i=1}^n c_i \bida_i = 0$
and $\sum_{i=1}^n c_i \bidb_i = 0$. We can view $\sigma$ inside the
lattice 
\begin{align*}
	\Sigma = \{ (d, c_1, \ldots, c_n) \in \bz_{\geq 0}^{n+1} : c_i \ge - d
\alpha_i \text{ for all }i \text{ and }\sum_{i=1}^n c_i \bida_i = \sum_{i =
1}^n c_i \bidb_i = 0 \}.	
\end{align*}

@@

For $i \in \Te(D)$, define \todo{below section needs to be simplified, maybe
by introducing new variables}
\begin{equation}\label{eqn:epsilon_i}
	\epsilon_i := (\prod_{k \ne i} f_k^{-\alpha_k \ell_i \gcd(\bida_i, \bidb_i)})
	(f_i^{\ell_i \frac{\gcd(\bida_i, \bidb_i)}{\bida_i + \bidb_i}\sum_{k \ne i}
	\alpha_k (\bida_k + \bidb_k)}) \in H^0(\ell_i \gcd(\bida_i, \bidb_i) D).
\end{equation}

\noindent
For $i \in \Tp(D)$ and $j \in \Tm(D)$, define
\begin{equation}\label{eqn:epsilon_i,j}
	\epsilon_{i, j} := \left(\prod_{k \ne i, j} f_k^{-\alpha_k \ell_{i,j} (\bida_i \bidb_
	j - \bida_j \bidb_i)} {f_i}^{s_1} {f_j}^{s_2}\right) \in H^0(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j \bidb_i) D
\end{equation}

@@

\noindent
and
\[
	s_2 = -\bidb_i \sum_{k \ne i, j} \alpha_k \bida_k + \bida_i \sum_{k \ne i, j}
	\alpha_k \bidb_k = -b_i (\alpha_i a_i + \alpha_j a_j) + a_i (\alpha_i b_i + \alpha_j b_j).
\]
\todo{Aaron: Why are $\epsilon_{i,j}$ defined here instead of $e_{i,j}$. We don't use the $\epsilon$ here, and we do use the $e$ here.}

The set of $e_i$'s and $e_{i,j}$'s corresponding to the $\epsilon_i$'s and $\epsilon_{i,j}$'s given above satisfies the conditions of
Proposition ~\ref{prop:cone-generation}, implying that $R_D$ is generated in 
degrees less than the sum of the degrees of the $e_i$ and $e_{i,j}$, which is
\[
	\sigma = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{\substack{
	i \in \Tp(D) \\	j \in \Tm(D)}} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\todo{Aaron: Do you use $\sigma$ later? We are usually reserving $\sigma$ for a
lattice point. Perhaps something like $\delta$ would be more appropriate if you
want to refer to it later. Also, you should number if if you want to refer to
it later. In general, it's bad style to define things in the proofs. This could
have been defined prior to the theorem statement as well, which would simplify
the statement. We should also then clarify that $\sigma$ depends on $D$.}

@@

Let $x_1, \ldots x_r$ be the generators of $R_D$ in degrees less than $\sigma$
(as given by Lemma ~\ref{lem:hirz-generators}), and let $\phi$ be the
surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through the
semigroup ring 
\[
	\bk[\Sigma] = \left \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d
	\alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i
	\right \rangle.
\]

@@

By Lemma ~\ref{lem:bound-ker-chi}, we can bound the degree of generation of $\ker(\chi)$ below
$2 \sigma$.
Finally, we calculate the degree of generation of $\psi$:

\begin{lem}
$\ker(\psi)$ is generated in less than
\[
	\tau = \sigma
	+ \max \left( \max_{i\in \Te}(\ell_i \gcd(a_i, b_i)), \max_{\substack{
	i \in	\Tp \\ j\in \Tm}} (\ell_{i,j} (\bida_i \bidb_j - \bida_j \bidb_i))
	\right).
\]
\end{lem}

\begin{proof}
We first claim that $\psi$ there are $\beta_1, \ldots, \beta_n$ \todo{Aaron:
Say where these polynomials live in}
 such that $\psi$ has relations of the form
\begin{equation}
\label{eqn:hirz-relations-psi}
	(z_i - \beta_i)\prod_{j=1}^n {z_j}^{c_{j}}
\end{equation}

\noindent
for any degree $d \in \mathbb{N}$ \todo{Aaron: This confuses me because $d$ is
not used above, is $d$ supposed to be $i$? Peter: Fixed?}
 and $c_i \ge -\alpha_i d$ satisfying $\bida_i + \sum_{j = 1}
^n \bida_j c_j = 0$ and $\bidb_i + \sum_{j=1}^n \bidb_j c_j = 0$.

To find such a $\beta$, \todo{Aaron: why is there no $i$ subscript? I changed
this in the $\bp^m$ case, probably copy past, mutatis mutandis}
 we simply write $f_i$ as a 
polynomial in $f_1, \ldots f_4$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{j =
1}^n f_j^{c_j}$ to the canonical form where $r_i = - \lfloor d \alpha_i
\rfloor$ whenever $i > 4$. \todo{Add $\ell_i$ defenition}

@@

Let $\sigma\in \Sigma'$ correspond to the monomial $y_i \prod_{j=1}^n
{y_j}^{c_j}$.  Then we can write $\sigma$ as a sum of $r_j e_j$'s for $j\in
\Te$ and $r_{j,k}e_{j,k}$ for $j\in \Tp, k\in \Tm$.  
For convenience, let $e_j$ (when it exists) be in degree $d_j$ with $j^{th}$
component $-\alpha_j d_j + \kappa_j$ and let $e_{i,j}$ (when it exists) lie in
degrees $d_{i,j}$ with $i^{th}$ component $\kappa_{i,j}'$ and $j^{th}$
component $\kappa_{j,i}''$.
Then a careful computation implies that 
%if $i\in \Te$, $i\in \Tp$, and $i\in \Tm$, then
\[
	\sigma = \lambda + E
\]
where $\lambda \in \Sigma$ and
\begin{equation}\label{eqn:hirz-E-translation}
	E = \begin{cases}
	\frac{1}{\kappa_i} e_i  	&  \mbox{ if }i\in \Te\\
	\sum_{j\in \Tm} r_{i,j} e_{i,j}: r_{i,j}\in \br, \sum_{j\in \Tm} r_{i,j}\kappa_{i,j}' = 1 & \mbox{ if } i\in \Tp\\
	\sum_{j\in \Tp} r_{j,i} e_{j,i}: r_{j,i} \in \br, \sum_{j\in \Tm} r_{j,i} \kappa_{i,j}'' = 1 	& \mbox{ if } i\in \Tm\\
	\end{cases}.
\end{equation}

\noindent
From this, we can deduce that
\[
	\deg(E) \le \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm \end{cases}.
\]
Let $M$ be the set of monomials terms of $\beta_j$.  Then for $\mu\in M$, we
can further compute that the lattice point corresponding to $\mu \prod_{j=1}^n
{f_i}^{c_i}$ is of the form
\[
	\sigma + E_\mu
\]
where $E_\mu$ is defined similarly to $E$ (in particular, they lie in the same degree).

Let $F\in R_D$ corresponding to the lattice point $E$ and let $F_\mu$ correspond to the lattice point $E_\mu$.  
Then $F - \sum_{\mu \in M} F_\mu = 0$ induces a
relation in degree at most
\[
	\sigma + \deg(F)
\]

@@

\noindent
Therefore, $\ker(\psi)$ is generated in degree at most
\[
	\tau = \sigma
	+ \max \left(\max_{i\in \Te} \bigl(\ell_i \gcd(a_i, b_i) \bigr),
	\; \max_{\substack{i \in \Tp \\ j \in \Tm}} \bigl(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j
	\bidb_i) \bigr) \right).
\]
\end{proof}

@@

\begin{thm}
\label{thm:hirz-generators-relations}
Let $D = \sum_{i=1}^n \alpha_i D_i \in \bq \otimes \di \hirz_k.$ Let $D_i =
V(f_i)$ where $f_i \in \sco(a_i, b_i)$; further suppose $f_1$ and $f_2$ are
independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$.
Then $R_D$ is generated in degrees at most
\[
	\sigma = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i +
	\sum_{\substack{i \in \Tp(D) \\	j \in \Tm(D)}} (\bida_i \bidb_j
	- \bida_j \bidb_i) \ell_{i, j}
\]

\noindent
with relations generated in degrees at most 
\[
	\max(2 \sigma, \sigma + \tau)
\]

\noindent
where
\[
	\tau = \sigma
	+ \max \left( \max_{i\in \Te} \bigl(\ell_i \gcd(a_i, b_i) \bigr),
	\max_{\substack{i \in \Tp \\ j \in \Tm}} \bigl(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j \bidb_i) \bigr) \right).
\]
\end{thm}
\todo{Aaron: Shouldn't $\tau$ not include $\sigma$?}

>>>>

\begin{lem}
\label{lem:hirz-generators}
$R_D$ is generated in degree at most
\begin{equation}\label{eqn:def-sigma}
	\rho = \sum_{i \in \Te(D)} \gcd(\bida_i, \bidb_i) \ell_i + \sum_{\substack
	{i \in \Tp(D) \\ j\in \Tm(D)}} (\bida_i \bidb_j - \bida_j \bidb_i)
	\ell	_{i,j}.
\end{equation}
\end{lem}

@@

\noindent
for some $c_i \ge - \alpha_i d$ such that $\sum_{i=1}^n c_i \bida_i = 0$
and $\sum_{i=1}^n c_i \bidb_i = 0$. We can view $g$ inside the
lattice 
\begin{align*}
	\Sigma = \{ (d, c_1, \ldots, c_n) \in \bz_{\geq 0}^{n+1} : c_i \ge - d
\alpha_i \text{ for all }i \text{ and }\sum_{i=1}^n c_i \bida_i = \sum_{i =
1}^n c_i \bidb_i = 0 \}.	
\end{align*}

@@

For $i \in \Te(D)$, define \todo{below section needs to be simplified, maybe
by introducing new variables}
\begin{equation}\label{eqn:epsilon_i}
	\epsilon_i := \left(\prod_{k \ne i} f_k^{-\alpha_k \ell_i \gcd(\bida_i, \bidb_i)}\right)
	(f_i^{\ell_i \frac{\gcd(\bida_i, \bidb_i)}{\bida_i + \bidb_i}\sum_{k \ne i}
	\alpha_k (\bida_k + \bidb_k)}) \in H^0(\ell_i \gcd(\bida_i, \bidb_i) D).
\end{equation}

\noindent
For $i \in \Tp(D)$ and $j \in \Tm(D)$, define
\begin{equation}\label{eqn:epsilon_i,j}
	\epsilon_{i, j} := \left(\prod_{k \ne i, j} f_k^{-\alpha_k \ell_{i,j} (\bida_i \bidb_
	j - \bida_j \bidb_i)} \right){f_i}^{s_1} {f_j}^{s_2} \in H^0(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j \bidb_i) D)
\end{equation}

@@

\noindent
and
\[
	s_2 = -\bidb_i \sum_{k \ne i, j} \alpha_k \bida_k + \bida_i \sum_{k \ne i, j}
	\alpha_k \bidb_k = -b_i (\alpha_i a_i + \alpha_j a_j) + a_i (\alpha_i b_i + \alpha_j b_j).
\]
\todo{Aaron: Why are $\epsilon_{i,j}$ defined here instead of $e_{i,j}$. We don't use the $\epsilon$ here, and we do use the $e$ here.}
A quick calculation deduces that these $\epsilon_i$'s and $\epsilon_{i,j}$'s lie on the extremal rays given by the $e_i$'s and $e_{i,j}$'s.

Thus Proposition ~\ref{prop:cone-generation} implies that $R_D$ is generated in 
degrees less than the sum of the degrees of the $\epsilon_i$'s and $\epsilon_{i,j}$'s, which is
\[
	\rho = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{\substack{
	i \in \Tp(D) \\	j \in \Tm(D)}} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\todo{Aaron: Do you use $\sigma$ later? We are usually reserving $\sigma$ for a
lattice point. Perhaps something like $\delta$ would be more appropriate if you
want to refer to it later. Also, you should number if if you want to refer to
it later. In general, it's bad style to define things in the proofs. This could
have been defined prior to the theorem statement as well, which would simplify
the statement. We should also then clarify that $\sigma$ depends on $D$.  \\
Peter: $\sigma$ replaced by $\rho$. It's also defined now in the lemma statement rather than its proof.  
}

@@

Let $x_1, \ldots x_r$ be the generators of $R_D$ in degrees less than $\rho$
(as given by Lemma ~\ref{lem:hirz-generators}), and let $\phi$ be the
surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through the
semigroup ring 
\[
	\bk[\Sigma] = \left \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d
	\alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i
	\right \rangle.
\]

@@

By Lemma ~\ref{lem:bound-ker-chi}, we can bound the degree of generation of $\ker(\chi)$ below
$2 \rho$.
Finally, we calculate the degree of generation of $\psi$:

\begin{lem}
$\ker(\psi)$ is generated in less than
\[
	\tau : = \rho
	+ \max \left(\max_{i\in \Te}(\ell_i \gcd(a_i, b_i)), \max_{\substack{
	i \in	\Tp \\ j\in \Tm}} (\ell_{i,j} (\bida_i \bidb_j - \bida_j \bidb_i))
	\right).
\]
\end{lem}

\begin{proof}
We first claim that $\psi$ there are $\beta_1, \ldots, \beta_n\in \bk[z_1, z_2, z_3, z_4]$
 such that $\psi$ has relations of the form
\begin{equation}
\label{eqn:hirz-relations-psi}
	(z_i - \beta_i)\prod_{j=1}^n {z_j}^{c_{j}}
\end{equation}

\noindent
whenever there is some degree $d \in \mathbb{N}$
and $c_i \ge -\alpha_i d$ satisfying $\bida_i + \sum_{j = 1}
^n \bida_j c_j = 0$ and $\bidb_i + \sum_{j=1}^n \bidb_j c_j = 0$.

To find such a $\beta_i$,
we simply write $f_i$ as a 
polynomial in $f_1, \ldots f_4$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{j =
1}^n f_j^{c_j}$ to the canonical form where $r_i = - \lfloor d \alpha_i
\rfloor$ whenever $i > 4$. \todo{Add $\ell_i$ defenition}

@@

Let $\sigma\in \Sigma'$ correspond to the monomial $y_i \prod_{j=1}^n
{y_j}^{c_j}$.  Then we can write $\sigma$ as a sum of $r_j e_j$'s for $j\in
\Te$ and $r_{j,k}e_{j,k}$ for $j\in \Tp, k\in \Tm$.  
For convenience, let $e_j$ (when it exists) be in degree $d_j$ with $j^{th}$
component $-\alpha_j d_j + \kappa_j$ for some $\kappa_j \in \br$ and let $e_{j,k}$ (when it exists) lie in
degrees $d_{j,k}$ with $j^{th}$ component $-\alpha_j d_{j,k} + \kappa_{j,k}'$ and $k^{th}$
component $-\alpha_j d_{j,k} + \kappa_{j,k}''$ for $\kappa_{j,k}', \kappa_{j,k}'' \in \br$
Then a careful computation deduces that 
%if $i\in \Te$, $i\in \Tp$, and $i\in \Tm$, then
\begin{equation}\label{eqn:sigma-in-terms-of-lambda/E}
	\sigma = \lambda + E
\end{equation}
where $\lambda \in \Sigma$ and
\begin{equation}\label{eqn:hirz-E-translation}
	E = \begin{cases}
	\frac{1}{\kappa_i} e_i  	&  \mbox{ if }i\in \Te\\
	\sum_{j\in \Tm} r_{i,j} e_{i,j}: r_{i,j}\in \br, \sum_{j\in \Tm} r_{i,j}\kappa_{i,j}' = 1 & \mbox{ if } i\in \Tp\\
	\sum_{j\in \Tp} r_{j,i} e_{j,i}: r_{j,i} \in \br, \sum_{j\in \Tm} r_{j,i} \kappa_{i,j}'' = 1 	& \mbox{ if } i\in \Tm\\
	\end{cases}.
\end{equation}

\noindent
From this, we can deduce that
\[
	\deg(E) \le \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm \end{cases}.
\]
Let $M$ be the set of monomials terms of $\beta_i$.  Then for $\mu\in M$, we
can further compute that the lattice point, $\sigma_\mu$, corresponding to $\mu \prod_{j=1}^n
{f_i}^{c_i}$ is of the form
\[
	\sigma_\mu = \lambda + E_\mu
\]
where $\lambda \in \Sigma$ is the same as in Equation \ref{eqn:sigma-in-terms-of-lambda/E} and $E_\mu$ is defined in a similar fashion as $E$ in Equation \ref{eqn:hirz-E-translation} (in particular, they lie in the same degree).

Let $F\in R_D$ corresponding to the lattice point $E$ and let $F_\mu$ correspond to the lattice point $E_\mu$.  
Then $F - \sum_{\mu \in M} F_\mu = 0$ induces a
relation in degree at most
\[
	\rho + \deg(F)
\]

@@

\noindent
Therefore, $\ker(\psi)$ is generated in degree at most
\[
	\tau = \rho
	+ \max \left(\max_{i\in \Te} \bigl(\ell_i \gcd(a_i, b_i) \bigr),
	\; \max_{\substack{i \in \Tp \\ j \in \Tm}} \bigl(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j
	\bidb_i) \bigr) \right).
\]
\end{proof}

@@

\begin{thm}
\label{thm:hirz-generators-relations}
Let $D = \sum_{i=1}^n \alpha_i D_i \in \bq \otimes \di \hirz_k.$ Let $D_i =
V(f_i)$ where $f_i \in \sco(a_i, b_i)$; further suppose $f_1$ and $f_2$ are
independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$.
Then $R_D$ is generated in degrees at most
\[
	\rho = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i +
	\sum_{\substack{i \in \Tp(D) \\	j \in \Tm(D)}} (\bida_i \bidb_j
	- \bida_j \bidb_i) \ell_{i, j}
\]

\noindent
with relations generated in degrees at most 
\[
	\max(2 \rho, \tau)
\]

\noindent
where
\[
	\tau = \sigma
	+ \max \left( \max_{i\in \Te} \bigl(\ell_i \gcd(a_i, b_i) \bigr),
	\max_{\substack{i \in \Tp \\ j \in \Tm}} \bigl(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j \bidb_i) \bigr) \right).
\]
\end{thm}
\todo{Aaron: Shouldn't $\tau$ not include $\sigma$?  Peter: Fixed}
